---
title: "Bioinformatics Class 11"
author: "Barry Grant"
date: "5/8/2018"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## PDB statistics

Downlad CSV file from PDB database <http://www.rcsb.org/stats/summary>. Read this into R and determine fraction of X-ray structures.

```{r}
pdb.stats <- read.csv("Data Export Summary.csv")
```

Lets claculate something

```{r}
precent <- (pdb.stats$Total / sum(pdb.stats$Total) ) * 100
names(precent) <- pdb.stats$Experimental.Method
precent
```


# Working with Bio3D

Lets load the package

```{r}
#install.packages("bio3d")
library(bio3d)
```

List the functions in the package and try an example...

```{r}
#lbio3d()
# example(plot.bio3d)
```

Read a PDB file

```{r}
pdb <- read.pdb("1hsg")
pdb
```

Select and write-out C-alpha atoms only!

```{r}
inds <- atom.select(pdb, "calpha")
inds
```

check if we have what we want

```{r}
head(pdb$atom[inds$atom,])
```

Looks good so lets continue to trim our PDB and write it out

```{r}
ca.pdb <- trim.pdb(pdb, inds=inds)
ca.pdb
write.pdb(ca.pdb, file="calpha_only.pdb")
```


## PCA of Adenylate kinase structures

Find the sequence

```{r}
aa <- get.seq("1ake_A")
aa
```


Search the PDB database

```{r, cache=TRUE}
b <- blast.pdb(aa)
```



```{r}
hits <- plot(b)
```


```{r}
head(hits$pdb.id)
```


```{r, cache=TRUE}
# Fetch PDBs
files <- get.pdb(hits$pdb.id, path = "pdbs", split = TRUE, gzip = TRUE)
```


```{r}
# Align structures
pdbs <- pdbaln(files)

# Vector containing PDB codes
ids <- basename.pdb(pdbs$id)

# Draw schematic alignment
plot(pdbs, labels=ids)
```





## PCA

```{r}
# find invariant core
core <- core.find(pdbs)

# superimpose all structures to core(
pdbs$xyz = pdbfit(pdbs, core)

# Perform PCA
pc.xray <- pca(pdbs)
```


```{r}
# Calculate RMSD
rd <- rmsd(pdbs)

# Structure-based clustering 
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd, k=3)
plot(pc.xray, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)

```


```{r}
 # Visualize first principal component 
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```

Tryout the brand new interactive 3D view functions!

```{r}
rgl.img <-  view(pc1)

rgl::rglwidget(elementId = "rgl.img")
```


View the fitted PDB

```{r}
rgl.pdbs <- view(pdbs)
rgl::rglwidget(elementId = "rgl.pdbs")
```





