---
title: "Bioinformatics Lecture 9"
author: "Barry Grant"
date: "5/1/2018"
output: 
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Hands-on with unsupervised learning analysis of cancer cells

Lets read our input cancer cell data

```{r}
url <- "https://bioboot.github.io/bimm143_W18/class-material/WisconsinCancer.csv"

# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(url)
head(wisc.df)
```

How many M's do we have
```{r}
table(wisc.df$diagnosis)
```


Make a numeric classification vector

```{r}
diagnosis <- as.numeric(wisc.df$diagnosis == "M")
table(diagnosis)
```

```{r}
# Convert the features of the data: wisc.data
wisc.data <- as.matrix( wisc.df[,3:ncol(wisc.df)] )
rownames(wisc.data) <- wisc.df$id
#head(wisc.data)
wisc.data <- wisc.data[,-31]
head(wisc.data)
```

## Exploratory data analysis

> The functions dim(), length(), grep() and sum() may be useful for answering the first 3 questions above.

**Q1**. How many observations are in this dataset?  

```{r}
nrow(wisc.data)
```

**Q2**. How many variables/features in the data are suffixed with _mean?

```{r}
# return things with the word `mean` in them
grep("mean", colnames(wisc.data), value = TRUE)

# return the position (ie index) of things with `mean` in them
grep("mean", colnames(wisc.data), value = FALSE) # default

# return things that DON'T have the word `mean` in them
grep("mean", colnames(wisc.data), value = TRUE, invert = TRUE)

length( grep("mean", colnames(wisc.data)) )
```


**Q3**. How many of the observations have a malignant diagnosis?

```{r}
sum(diagnosis)
```


# Section 2. Performing PCA

Check if we need to `scale=TRUE` our data

```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

Demo the minus use for extracting positions/cols/rows etc.

```{r}
x <- c("barry", "chris", "mary", "chandra")
x[-c(1:3)]
```

Now do PCA with scalling

```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data, scale=TRUE)
```

Inspect a summary of the results with the summary() function.

```{r}
summary(wisc.pr)

```

## Plot PCA results

What is in our PCA output
```{r}
attributes(wisc.pr)
```


```{r}
# Scatter plot observations by components 1 and 2
plot( wisc.pr$x[,1], wisc.pr$x[,2], col = diagnosis+1,
      pch=as.vector(wisc.df$diagnosis), 
     xlab = "PC1", ylab = "PC2")
```

## Varience Explained: Scree-plot

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2

# Proportion of varance
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")

```

plot as a barplot

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)

axis(2, at=pve, labels=round(pve,2)*100 )
```

Combined plots

```{r}
par(mfrow=c(1,2))

plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")

plot(cumsum(pve), xlab = "Principal Component", 
     ylab = "Cumulative Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")

```

# Hierarchical clustering

Lets do hierarchical clustering ;-)

```{r}
#?hclust
#?dist

# Scale the wisc.data data: data.scaled
data.scaled <- scale(wisc.data)

# Calculate distance matrix needed for hclust
data.dist <- dist(data.scaled)

# Clustering by hclust()
wisc.hclust <- hclust(data.dist)
                      
```

Plot our hclust model tree

```{r}
plot(wisc.hclust)
abline(h=20, col="red")

# Cut into 4 clusters/groups
wisc.hclust.clusters <- cutree(wisc.hclust, h=20)

# Or:
# wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
```

```{r}
table(wisc.hclust.clusters)
```


How may M or 1 (cancer) and 0 (non-cancer) are in each cluster
```{r}
table(wisc.hclust.clusters, diagnosis)
```



## K-means

```{r}
wisc.km <- kmeans(scale(wisc.data), 2, nstart = 20)
table(wisc.km$cluster, diagnosis)
```



Compare kmeans with hclust

```{r}
table(wisc.hclust.clusters, wisc.km$cluster)
```



## Cluster in PC space

```{r}
## Use the distance along the first 7 PCs for clustering i.e. wisc.pr$x[, 1:7]
wisc.pr.hclust <- hclust( dist(wisc.pr$x[, 1:7]), method="ward.D2" )
plot(wisc.pr.hclust)
```


Cut this hierarchical clustering model into 2 clusters and assign the results to `wisc.pr.hclust.clusters`.

```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
```

Using `table()`, compare the results from your new hierarchical clustering model with the actual diagnoses. 

```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```

Lets plot with our clustering results

```{r}
# Scatter plot observations by components 1 and 2
plot( wisc.pr$x[,1:2], col = wisc.pr.hclust.clusters,
     xlab = "PC1", ylab = "PC2")
```








