---
title: "Bioinformatics Class 8"
author: "Barry Grant"
date: "4/26/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# K-means ckustering

Lets make up some data to try K-means on:

```{r}
# Generate some example data for clustering
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))

plot(x)
```

Our tasks
 - Use the kmeans() function setting k to 2 and nstart=20 

```{r}
km <- kmeans(x, centers = 2, nstart=20)
km
```

```{r}
km$cluster
```



Inspect/print the results 

Q. How many points are in each cluster?

```{r}
km$size
```

Q. What ‘component’ of your result object details 
      - cluster size?
      - cluster assignment/membership? 
      - cluster center?

```{r}
# cluster assignment/membership? 
km$cluster
```

```{r}
# cluster center
km$centers
```



Plot x colored by the kmeans cluster assignment and 
      add cluster centers as blue points

```{r}
plot(x, col=km$cluster)
points(km$centers, col="blue", pch=15)
```
      
      
      
 Q. Repeat for k=3, which has the lower tot.withinss? 
```{r}
km3 <- kmeans(x, centers = 3, nstart=20)
km3
```

```{r}
km3$tot.withinss
```


# Hierarchical clustering

Lets try on our x points

```{r}
# First we need to calculate point (dis)similarity
#   as the Euclidean distance between observations 
dist_matrix <- dist(x) 
```

Lets look at this distance matrix

```{r}
class(dist_matrix)
```


```{r}
dim(as.matrix(dist_matrix))

```



```{r}
# The hclust() function returns a hierarchical 
#  clustering model
hc <- hclust(d = dist_matrix)

# the print method is not so useful here 
hc  
```

Lets draw the tree

```{r}
plot(hc)
```

I can 'cut' the tree at any height to give our clusters...

```{r}
# Cut at height 6
cutree(hc, h=6)
```

Or ask it to cut yieling `k` clusters

```{r}
# Give me 4 clusters please!
cutree(hc, k=4)
```

## Another example

Make up some data

```{r}
# Step 1. Generate some example data for clustering
x <- rbind(
  matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2),   # c1
  matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
  matrix(c(rnorm(50, mean = 1, sd = 0.3),           # c3
           rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")

# Step 2. Plot the data without clustering
plot(x)

# Step 3. Generate colors for known clusters 
#         (just so we can compare to hclust results)
col <- as.factor( rep(c("c1","c2","c3"), each=50) ) 

plot(x, col=col)

```


**Q.** Use the dist(), hclust(), plot() and cutree() functions to return 2 and 3 clusters 

```{r}
# Distance matrix
d <- dist(x)
hc <- hclust(d)
plot(hc)
```

```{r}
# To return 2 clusters I could cut at h=2.5 or set k=2
members2 <- cutree(hc, k=2)

# To return 3 clusters I could cut at h=2 or set k=3
members3 <- cutree(hc, k=3)

# lets look at both
members <- cbind(k2=members2, k3=members3)
```

Plot
```{r}
plot(x, col=members3, pch=15)
```


**Q.** How does this compare to your known 'col' groups?

## PCA :-)

Lets make up some data

```{r}
## Initialize a blank 100 row by 10 column matrix
mydata <- matrix(nrow=100, ncol=10)
dim(mydata)
```

```{r}
## Lets label the rows gene1, gene2 etc. to gene100
rownames(mydata) <- paste("gene", 1:100, sep="")
```

```{r}
## Lets label the first 5 columns wt1, wt2, wt3, wt4 and wt5
##   and the last 5 ko1, ko2 etc. to ko5 (for "knock-out") 
colnames(mydata) <- c(paste("wt", 1:5, sep=""),
                    paste("ko", 1:5, sep="") )

## Fill in some fake read counts
for(i in 1:nrow(mydata)) {
  wt.values <- rpois(5, lambda=sample(x=10:1000, size=1))
  ko.values <- rpois(5, lambda=sample(x=10:1000, size=1))

  mydata[i,] <- c(wt.values, ko.values)
}

head(mydata)
```

```{r}
#head(t(mydata))
```

```{r}
pca <- prcomp( t(mydata), scale=TRUE )
pca
```


Lets plot our PCA plot
```{r}
## A basic PC1 vs PC2 2-D plot
plot(pca$x[,1], pca$x[,2])

```


How well are our PC's capturing our data spread (i.e. variance)

```{r}
## Precent variance is often more informative to look at 
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)

pca.var.per

```


Plot our scree plot 

```{r}
barplot(pca.var.per, main="Scree Plot", 
        xlab="Principal Component", ylab="Percent Variation")
```

```{r}
# Color up our PCA plot
colvec <- as.factor( substr( colnames(mydata), start=1, stop=2) )

plot(pca$x[,1], pca$x[,2], col=colvec)

```



# UK foods

```{r}
# Read data
 x <- read.csv("https://bioboot.github.io/bggn213_S18/class-material//UK_foods.csv", row.names = 1)

```

Plot things

```{r}
par(mfrow=c(4,1), mar=c(2,4,0,2), las=2)
ymax <- max(x)
barplot(x[,1], ylim=c(0,ymax))
barplot(x[,2], ylim=c(0,ymax))
barplot(x[,3], ylim=c(0,ymax))
par(mar=c(4,4,0,1))
barplot(x[,4], names.arg=rownames(x), las=2, ylim=c(0,ymax))
```

# Do heatmap clustering and  PCA

See class website: < https://bioboot.github.io/bggn213_S18/class-material/UK_food_pca/ >


