---
title: "Bioinformatics Class7"
author: "Barry Grant"
date: "4/24/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Functions again

Here I am going to revist our function from class 6

```{r}
source("http://tinyurl.com/rescale-R")
ls()
```

Lets see if we can use this function

```{r}
rescale(1:10)
```

Looks good lets break it!

```{r, eval=FALSE}
rescale( c(1:10,"string") )
```

Lets try the new **rescale2()** function

```{r}
rescale2( c(1:10,"string") )
```

## Write a NA checking function

Here we write a new function to check for NAs in two inputs

```{r}
x <- c( 1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3,  4)
```


```{r}
which( is.na(x) )

```

```{r}
sum( is.na(x) )
```

Lets put this together with `x` and `y`

```{r}
is.na(x)
is.na(y)
```

```{r}
is.na(x) & is.na(y)
```

```{r}
# Putting together!
sum( is.na(x) & is.na(y) )

```

This is the both_na() function starting point

```{r}
both_na <- function(x, y) {
  ## Check for NA elements in both input vectors 
  sum( is.na(x) & is.na(y) )
}

```


Test it 

```{r}
both_na(x,y)
```

and again

```{r}
x <-  c(NA, NA, NA)
y1 <- c( 1, NA, NA)
y2 <- c( 1, NA, NA, NA)

both_na(x, y2)
```

Use stop() inside function to catch this case

```{r, eval=FALSE}
both_na2(x, y2)
```

Try the both_na3() function

```{r}
x <- c( 1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3,  4)

ans <- both_na3(x, y)
```

```{r}
ans$number
```

## Anoter function example: gene intersection

```{r}
df1
df2

x <- df1$IDs
y <- df2$IDs
```

```{r}
x
y
```


```{r}
#intersect(x, y)

x %in% y
```

```{r}
y %in% x
```


Now we can access the genes we want with these 'indices'

```{r}
x[x %in% y]
y[y %in% x]

```

Can make these columns of the same object with **cbind()**

```{r}
cbind( x[x %in% y], y[y %in% x])
```

```{r}
# Correct but yucky format for 2nd colnames
gene_intersect2(df1, df2)

```

Lets try the merge() function for this job

```{r}
merge(df1, df2, by="IDs")

```





